Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 paper.qmd
True model equation (removed gamma)
Updated Figure 7 text

To illustrate the method, we use $7\text{-}D$ simulated data, which we call the "nonlinear clusters". It is constructed by simulating two clusters, each consisting of $1000$ observations. The C-shaped cluster is generated from $\theta \sim U(\text{-}3\pi/2, 0)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the C), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$, $X_4 = \cos(\theta)$. The other cluster is from $X_1 \sim U(0, 2)$, $X_2 \sim U(0, 3)$, $\gamma \sim U(0, 0.5)$, $X_3 = \text{-}(X_1^3 + X_2) + \gamma$, and $X_4 \sim U(0, 2)$. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

mse_two_curvy_b <- ggplot(error_two_curvy_tsne, 
                     aes(x = a1, 
                         y = RMSE)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "RMSE") +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

prop_dens_a1 <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  interior_annotation("d", position = c(0.9, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))

a1_m_two_curvy <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("b", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

@@

mse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = RMSE,
                         colour = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 3, 5, 7, 9, 11, 13, 15, 17)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted"),
                        labels = parse_format()) +
  xlab("bin density cutoff") +
  ylab("RMSE") +
  interior_annotation("c", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best hexagon bin parameters (a, b, d) and thresholds to remove low-density bins (c). Both (a) and (c) show RMSE, against binwidth ($a_1$) and bin density cutoff. A good benchmark value for these parameters is when the RMSE drops and then flattens out. Three binwidth choices were made: $0.03$ (orange dashed), $0.05$ (blue solid), and $0.09$ (green dotted) to investigate. As the binwidth increases, the proportion of non-empty bins also increases (b). The relative proportion density decreases and levels off (d). Binwidth $0.05$ is chosen as the initial best binwidth for further analysis. There is no need to remove the low-density hexagons because as shown in (b), there is no considerable drop in RMSE."
#| label: fig-param-two-curvy
#| out-width: 80%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

mse_two_curvy_b + a1_m_two_curvy +
  mse_two_curvy_lwd + prop_dens_a1 +
  plot_layout(ncol=2, guides = "collect")
```

>>>>

To illustrate the method, we use $7\text{-}D$ simulated data, which we call the "nonlinear clusters". It is constructed by simulating two clusters, each consisting of $1000$ observations. The C-shaped cluster is generated from $\theta \sim U(\text{-}3\pi/2, 0)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the C), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$, $X_4 = \cos(\theta)$. The other cluster is from $X_1 \sim U(0, 2)$, $X_2 \sim U(0, 3)$, $X_3 = \text{-}(X_1^3 + X_2)$, and $X_4 \sim U(0, 2)$. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

mse_two_curvy_b <- ggplot(error_two_curvy_tsne, 
                     aes(x = a1, 
                         y = RMSE)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "RMSE") +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

prop_dens_a1 <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("relative ", bar(d[h])))) +
  interior_annotation("d", position = c(0.9, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))

a1_m_two_curvy <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("b", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

@@

mse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = RMSE,
                         colour = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 3, 5, 7, 9, 11, 13, 15, 17)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted"),
                        labels = parse_format()) +
  xlab("bin density cutoff") +
  ylab("RMSE") +
  interior_annotation("c", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best hexagon bin parameters (a, b, d) and thresholds to remove low-density bins (c). Both (a) and (c) show RMSE, against binwidth ($a_1$) and bin density cutoff. A good benchmark value for these parameters is when the RMSE drops and then flattens out. Three binwidth choices were made: $0.03$ (orange dashed), $0.05$ (blue solid), and $0.08$ (green dotted) to investigate. As the binwidth increases, the proportion of non-empty bins also increases (b). The relative $\\bar{d_h}$ decreases and levels off (d). Binwidth $0.05$ is chosen as the initial best binwidth for further analysis. There is no need to remove the low-density hexagons because as shown in (b), there is no considerable drop in RMSE."
#| label: fig-param-two-curvy
#| out-width: 80%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

mse_two_curvy_b + a1_m_two_curvy +
  mse_two_curvy_lwd + prop_dens_a1 +
  plot_layout(ncol=2)
```
